### 2022-06-02 chapters/fairness/fif/experiments.tex
minor edits in many parts of the thesis

\noindent\textbf{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $). In {\fairXplainer}, we deploy SAlib library~\cite{Herman2017} to compute FIFs leveraging techniques of global sensitivity analysis. In our experiments, we consider five widely studied datasets from fairness literature, namely COMPAS~\cite{angwin2016machine}, Adult~\cite{DK2017uci}, German-credit~\cite{DK2017}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}). We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: Logistic Regression, Support Vector Machine, Decision Tree, and Neural Networks. In experiments, we specify {\fairXplainer} to compute intersectional influences up to second order ($ \lambda = 2 $). We compare {\fairXplainer} with Shapley-valued based FIF computational framework, referred as SHAP~\cite{lundberg2020explaining}. In addition, we deploy {\fairXplainer} along with different fairness-enhancing~\cite{kamiran2012data} and fairness attack~\cite{solans2020poisoning} algorithms, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objective of our empirical study. 
\begin{enumerate}[nosep,leftmargin=*]
%\begin{enumerate}
	\item How do \textbf{accuracies} of {\fairXplainer} and SHAP compare for estimating FIFs and resulting bias?
	\item What is the impact of\textbf{ intersectional vs.\ individual FIFs} in tracing the sources of bias?
    \item How do FIFs change while \textbf{applying} different fairness enhancing algorithms, i.e. \textbf{affirmative actions}, and fairness attacks, i.e. \textbf{punitive actions}?
	%\item \red{What is the effect of $ \lambda $ on the runtime of {\fairXplainer}?}
\end{enumerate}
In summary, \textit{{\fairXplainer} achieves significantly less estimation error than SHAP in all the datasets.} This shows the importance of adopting global sensitivity analysis to compute FIFs for group fairness metrics than local explanation approaches. Moreover, \textit{{\fairXplainer} effectively traces the sources of bias by computing intersectional FIFs, which earlier method like SHAP {did not capture}.} {\fairXplainer} also detects the effects of the affirmative and punitive actions on the bias of a classifier and the corresponding tensions between different subsets of features. %Due to brevity of space, we defer additional experimental results to the Supplementary material. %In the following, we elaborate results on estimation error and individual vs.\ intersectional FIFs.

\begin{figure}[!t]
	\begin{minipage}{0.49\textwidth}
		\centering
		\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/fairness/fif/sp_train_accuracy}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/fairness/fif/eo_train_accuracy}}
	\end{minipage}
	\caption[Accuracy of {\fairXplainer}]{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics. Lower values on the $ Y $-axis denote a better result. {\fairXplainer} has significantly less error than SHAP.}\label{fig:estimation_error}
%	\vspace{-0.5em}
\end{figure}
\noindent\textbf{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}). We consider five datasets, each trained on different classifiers by applying five-fold cross-validation. We compute estimation error by taking the absolute difference between the exact and  estimated value of a metric, and present results in Figure~\ref{fig:estimation_error}. In both metrics, {\fairXplainer} demonstrates significantly less error than SHAP. For example, the mean error of {\fairXplainer} in estimating $ \mathsf{SP} $ on Titanic dataset is $ 0.1 $ vs.\ $ 0.42 $ for SHAP. In this context, the error of {\fairXplainer} is often zero in most datasets and only insignificant ($ \le 0.13 $) due to the degenerate cases (Section~\ref{sec:fifs}) with $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{0, 1\} $.  Therefore, \emph{global sensitivity analysis based approach {\fairXplainer} is significantly more accurate in computing FIFs and corresponding fairness metric than local explanation based approach SHAP}.

@@

\noindent\textbf{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs. We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity as $ 0.17 $, which means a non-Caucasian  has $ 0.17 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the source of bias and present individual FIFs in Figure~\ref{fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fig:individual_and_intersectional_fifs}. In both figures, we show top seven influential FIFs sorted by absolute values,  followed by residual higher-order FIFs. In Figure~\ref{fig:individual_fifs}, `priors count' of an individual is dominating in increasing statistical parity (FIF $ = 0.12 $). Other non-sensitive features appear to have almost zero FIFs. However, the sum of second-order influences of all features increases statistical parity by $ 0.06 $, denoting that the data is highly correlated and presenting only individual FIFs does not trace the true sources of bias. For example, while both `sex' and `age' of persons have zero influence on bias (Figure~\ref{fig:individual_fifs}), these features together with `c charge degree', `priors count', and `juvenile other count' contribute highly on statistical parity (sum of FIFs $ = 0.03 + 0.02 + 0.01 + 0.01 - 0.01 - 0.01 = 0.05$). \emph{Therefore, intersectional influences together with individual influences lead to a clearer understanding on the source of bias of a classifier. Note that, unlike SHAP, {\fairXplainer} is the only framework that computes beyond individual FIFs.}
\begin{figure}[t!]
	\begin{minipage}{0.43\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fig:individual_fifs}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fig:individual_and_intersectional_fifs}}
	\end{minipage}
	\vspace{-0.2em}
	\caption[Individual vs.\ intersectional FIFs]{FIFs for COMPAS dataset on explaining statistical parity. Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}

@@

\noindent\textbf{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.  In Figure~\ref{fig:individual_vs_intersectional_influence}, statistical parity of the classifier is $ 0.17 $. Applying a data poisoning fairness attack~\cite{solans2020poisoning} increases statistical parity to $ 0.5 $ and the data reweighing-based fairness-enhancing algorithm~\cite{kamiran2012data} decreases it to $ 0.15$. We observe that in both cases, there are a subset of features decreasing bias while another subset of features increasing it, e.g. `age' vs.\ `priors count'. Figure~\ref{fig:fairness_attack} suggests that the attack algorithm would be more successful if it could hide the influence of `age' of a person in receiving discriminating prediction for re-offending crimes. Figure~\ref{fig:fairness_repair} suggests that the fairness enhancing algorithm can improve by ameliorating the effect of `priors count' further. Thus, \textit{{\fairXplainer} provides a dissecting tool to undertake necessary steps to improve or worsen fairness of the classifier.}

>>>>

\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $). In {\fairXplainer}, we deploy SAlib library~\cite{Herman2017} to compute FIFs leveraging techniques of global sensitivity analysis. In our experiments, we consider five widely studied datasets from fairness literature, namely COMPAS~\cite{angwin2016machine}, Adult~\cite{DK2017uci}, German-credit~\cite{DK2017}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}). We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: Logistic Regression, Support Vector Machine, Decision Tree, and Neural Networks. In experiments, we specify {\fairXplainer} to compute intersectional influences up to second order ($ \lambda = 2 $). We compare {\fairXplainer} with Shapley-valued based FIF computational framework, referred as SHAP~\cite{lundberg2020explaining}. In addition, we deploy {\fairXplainer} along with different fairness-enhancing~\cite{kamiran2012data} and fairness attack~\cite{solans2020poisoning} algorithms, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objective of our empirical study.

\begin{enumerate}
%\begin{enumerate}
	\item How do \textbf{accuracies} of {\fairXplainer} and SHAP compare for estimating FIFs and resulting bias?
	\item What is the impact of\textbf{ intersectional vs.\ individual FIFs} in tracing the sources of bias?
    \item How do FIFs change while \textbf{applying} different fairness enhancing algorithms, i.e. \textbf{affirmative actions}, and fairness attacks, i.e. \textbf{punitive actions}?
	%\item \red{What is the effect of $ \lambda $ on the runtime of {\fairXplainer}?}
\end{enumerate}

In summary, \textit{{\fairXplainer} achieves significantly less estimation error than SHAP in all the datasets.} This shows the importance of adopting global sensitivity analysis to compute FIFs for group fairness metrics than local explanation approaches. Moreover, \textit{{\fairXplainer} effectively traces the sources of bias by computing intersectional FIFs, which earlier method like SHAP {did not capture}.} {\fairXplainer} also detects the effects of the affirmative and punitive actions on the bias of a classifier and the corresponding tensions between different subsets of features. %Due to brevity of space, we defer additional experimental results to the Supplementary material. %In the following, we elaborate results on estimation error and individual vs.\ intersectional FIFs.

\begin{figure}[!t]
	\begin{minipage}{0.49\textwidth}
		\centering
		\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/fairness/fif/sp_train_accuracy}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/fairness/fif/eo_train_accuracy}}
	\end{minipage}
	\caption[Accuracy of {\fairXplainer}]{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics. Lower values on the $ Y $-axis denote a better result. {\fairXplainer} has significantly less error than SHAP.}\label{fig:estimation_error}
%	\vspace{-0.5em}
\end{figure}
\paragraph{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}). We consider five datasets, each trained on different classifiers by applying five-fold cross-validation. We compute estimation error by taking the absolute difference between the exact and  estimated value of a metric, and present results in Figure~\ref{fig:estimation_error}. In both metrics, {\fairXplainer} demonstrates significantly less error than SHAP. For example, the mean error of {\fairXplainer} in estimating $ \mathsf{SP} $ on Titanic dataset is $ 0.1 $ vs.\ $ 0.42 $ for SHAP. In this context, the error of {\fairXplainer} is often zero in most datasets and only insignificant ($ \le 0.13 $) due to the degenerate cases (Section~\ref{sec:fifs}) with $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{0, 1\} $.  Therefore, \emph{global sensitivity analysis based approach {\fairXplainer} is significantly more accurate in computing FIFs and corresponding fairness metric than local explanation based approach SHAP}.

@@

\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs. We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity as $ 0.17 $, which means a non-Caucasian  has $ 0.17 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the source of bias and present individual FIFs in Figure~\ref{fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fig:individual_and_intersectional_fifs}. In both figures, we show top seven influential FIFs sorted by absolute values,  followed by residual higher-order FIFs. In Figure~\ref{fig:individual_fifs}, `priors count' of an individual is dominating in increasing statistical parity (FIF $ = 0.12 $). Other non-sensitive features appear to have almost zero FIFs. However, the sum of second-order influences of all features increases statistical parity by $ 0.06 $, denoting that the data is highly correlated and presenting only individual FIFs does not trace the true sources of bias. For example, while both `sex' and `age' of persons have zero influence on bias (Figure~\ref{fig:individual_fifs}), these features together with `c charge degree', `priors count', and `juvenile other count' contribute highly on statistical parity (sum of FIFs $ = 0.03 + 0.02 + 0.01 + 0.01 - 0.01 - 0.01 = 0.05$). \emph{Therefore, intersectional influences together with individual influences lead to a clearer understanding on the source of bias of a classifier. Note that, unlike SHAP, {\fairXplainer} is the only framework that computes beyond individual FIFs.}
\begin{figure}[t!]
	\begin{minipage}{0.43\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fig:individual_fifs}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fig:individual_and_intersectional_fifs}}
	\end{minipage}
	\vspace{-0.2em}
	\caption[Individual vs.\ intersectional FIFs]{FIFs for COMPAS dataset on explaining statistical parity. Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}

@@

\paragraph{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.  In Figure~\ref{fig:individual_vs_intersectional_influence}, statistical parity of the classifier is $ 0.17 $. Applying a data poisoning fairness attack~\cite{solans2020poisoning} increases statistical parity to $ 0.5 $ and the data reweighing-based fairness-enhancing algorithm~\cite{kamiran2012data} decreases it to $ 0.15$. We observe that in both cases, there are a subset of features decreasing bias while another subset of features increasing it, e.g. `age' vs.\ `priors count'. Figure~\ref{fig:fairness_attack} suggests that the attack algorithm would be more successful if it could hide the influence of `age' of a person in receiving discriminating prediction for re-offending crimes. Figure~\ref{fig:fairness_repair} suggests that the fairness enhancing algorithm can improve by ameliorating the effect of `priors count' further. Thus, \textit{{\fairXplainer} provides a dissecting tool to undertake necessary steps to improve or worsen fairness of the classifier.}

### 2022-06-02 chapters/fairness/fif/fairxplainer.tex
minor edits in many parts of the thesis

\noindent\textbf{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}). Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \function_{\mathbf{a}}(\nonsensitive) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $. We express $ \function_{\mathbf{a}} $ as a set-additive model:
\begin{align}
\label{eq:set_additive_classifier}
\function_{\mathbf{a}}(\mathbf{X}) = \function_{\mathbf{a}, 0} +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} \function_{\mathbf{a},\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \delta
\end{align}
\begin{algorithm}
	\caption{{\fairXplainer}}\label{algo:framework}
	\hspace*{\algorithmicindent} \textbf{Input:} $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y},  \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n, f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}, \lambda $  \\
	\hspace*{\algorithmicindent} \textbf{Output:} $ w_{\mathbf{S}}$ 
	\begin{algorithmic}[1]
		\State $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], k \gets    |\nonsensitive| $
		\label{algo_line:fif_computation_start}

@@

\noindent\textbf{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}). Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \function_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed. Specifically, $ \function_{\mathbf{a}, \mathbf{S}} $ is learned as a smoothed function of $ g $ and rest of the components $ \function_{\mathbf{a}, \mathbf{S}'} $, where $ \mathbf{S}' \ne \mathbf{S} $ is a non-empty subset of $ [k]\setminus \emptyset $. To keep every component function mean centered, backfitting requires to impose the constraints $ \function_{\mathbf{a}, 0} = \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n) $ (Line~\ref{alg_line:initialization}), which is the mean of $ \function_{\mathbf{a}} $ evaluated on samples belonging to the sensitive group $ \sensitive = \mathbf{a} $;  and $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}) = 0$ (Line~\ref{alg_line:mean_centered}), where $ \mathbf{x}_{\mathbf{S}}^{(i)} $ be the subset of feature values associated with feature indices $ \mathbf{S} $ for the $ i $-th sample $ \mathbf{x}^{(i)} $. These constraints assign the expectation of $ \function_{\mathbf{a}} $ on the constant term $ \function_{\mathbf{a}, 0} $ and the variance of $ \function_{\mathbf{a}} $ to the component functions.

While performing local regression, backfitting uses a smoothing operator~\cite{loader2012smoothing} over the set of samples (Line~\ref{alg_line:backfitting_step}). A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating the local curves obtained by local regression~\cite{loader2012smoothing}. In this paper, we apply cubic spline smoothing~\cite{li2010global} to learn each component function. Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity. Hence, up to the second derivatives of each piecewise term are zero at the endpoints of intervals. We refer to Appendix~\ref{sec:smoothing} for further details.%\todo{Add pseudo-code of {\fairXplainer}}

@@

\noindent\textbf{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $. Since each component function is mean centered (Line~\ref{alg_line:mean_centered}), we compute the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ on the dataset as
$  \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n (\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}))^2 $. Hence, variance captures the independent effect of $ \function_{\mathbf{a}, \mathbf{S}} $.
%$ V_{\mathbf{a}, \mathbf{S}} $ is the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ evaluated on each $ \mathbf{x}_{\mathbf{S}}^{(i)} $ in the dataset. 
Covariance is computed to account for the correlation among features $ \nonsensitive $. We compute the covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ with $ \function_{\mathbf{a}} $ on the dataset as
$
\label{eq:covariance_computation}
	 \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}). 
$
Here, $ g_\mathbf{a}(\cdot) - \function_{\mathbf{a}, 0} $ is the mean centered form of $ \function_{\mathbf{a}} $. Covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ can be both positive and negative depending on whether the features $ \nonsensitive_{\mathbf{S}} $ are positively or negatively correlated with $ \function_{\mathbf{a}} $. Specifically, under the set additive model, we obtain $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $. Now, we use $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] $ as the effective variance of $ \nonsensitive_{\mathbf{S}} $ for a given sensitive group $ \sensitive = \mathbf{a} $ (Line~\ref{algo_line:covariance_computation_start}--\ref{algo_line:covariance_computation_end}). In Line~\ref{algo_line:fif_computation_start}--\ref{algo_line:fif_computation_end}, we compute $ V_{\mathbf{a}, \mathbf{S}} $ for the most and the least favored groups, and plug them in Theorem~\ref{lemma:fif} to compute FIF of $ \nonsensitive_{\mathbf{S}} $.

@@

\noindent\textbf{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
We extend {\fairXplainer} in computing FIF of group fairness metrics beyond statistical parity, namely equalized odds and predictive parity. For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of samples in the dataset where $ Y = 1 $ and another on samples with $ Y = 0 $. Then, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ quantifies the equalized odds of the classifer.

>>>>

\paragraph{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}). Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \function_{\mathbf{a}}(\nonsensitive) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $. We express $ \function_{\mathbf{a}} $ as a set-additive model:

\begin{align}
\label{eq:set_additive_classifier}
\function_{\mathbf{a}}(\mathbf{X}) = \function_{\mathbf{a}, 0} +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} \function_{\mathbf{a},\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \delta
\end{align}

\begin{algorithm}
	\caption{{\fairXplainer}}\label{algo:framework}
	\hspace*{\algorithmicindent} \textbf{Input:} $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y},  \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n, f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}, \lambda $  \\
	\hspace*{\algorithmicindent} \textbf{Output:} $ w_{\mathbf{S}}$ 
	\begin{algorithmic}[1]
		\State $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], k \gets    |\nonsensitive| $
		\label{algo_line:fif_computation_start}

@@

\paragraph{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}). Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \function_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed. Specifically, $ \function_{\mathbf{a}, \mathbf{S}} $ is learned as a smoothed function of $ g $ and rest of the components $ \function_{\mathbf{a}, \mathbf{S}'} $, where $ \mathbf{S}' \ne \mathbf{S} $ is a non-empty subset of $ [k]\setminus \emptyset $. To keep every component function mean centered, backfitting requires to impose the constraints $ \function_{\mathbf{a}, 0} = \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n) $ (Line~\ref{alg_line:initialization}), which is the mean of $ \function_{\mathbf{a}} $ evaluated on samples belonging to the sensitive group $ \sensitive = \mathbf{a} $;  and $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}) = 0$ (Line~\ref{alg_line:mean_centered}), where $ \mathbf{x}_{\mathbf{S}}^{(i)} $ be the subset of feature values associated with feature indices $ \mathbf{S} $ for the $ i $-th sample $ \mathbf{x}^{(i)} $. These constraints assign the expectation of $ \function_{\mathbf{a}} $ on the constant term $ \function_{\mathbf{a}, 0} $ and the variance of $ \function_{\mathbf{a}} $ to the component functions.

While performing local regression, backfitting uses a smoothing operator~\cite{loader2012smoothing} over the set of samples (Line~\ref{alg_line:backfitting_step}). A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating the local curves obtained by local regression~\cite{loader2012smoothing}. In this chapter, we apply cubic spline smoothing~\cite{li2010global} to learn each component function. Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity. Hence, up to the second derivatives of each piecewise term are zero at the endpoints of intervals. We refer to Appendix~\ref{sec:smoothing} for further details.%\todo{Add pseudo-code of {\fairXplainer}}

@@

\paragraph{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $. Since each component function is mean centered (Line~\ref{alg_line:mean_centered}), we compute the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ on the dataset as
$  \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n (\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}))^2 $. Hence, variance captures the independent effect of $ \function_{\mathbf{a}, \mathbf{S}} $.
%$ V_{\mathbf{a}, \mathbf{S}} $ is the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ evaluated on each $ \mathbf{x}_{\mathbf{S}}^{(i)} $ in the dataset. 
Covariance is computed to account for the correlation among features $ \nonsensitive $. We compute the covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ with $ \function_{\mathbf{a}} $ on the dataset as
$
\label{eq:covariance_computation}
	 \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}). 
$
Here, $ g_\mathbf{a}(\cdot) - \function_{\mathbf{a}, 0} $ is the mean centered form of $ \function_{\mathbf{a}} $. Covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ can be both positive and negative depending on whether the features $ \nonsensitive_{\mathbf{S}} $ are positively or negatively correlated with $ \function_{\mathbf{a}} $. Specifically, under the set additive model, we obtain $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $. Now, we use $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] $ as the effective variance of $ \nonsensitive_{\mathbf{S}} $ for a given sensitive group $ \sensitive = \mathbf{a} $ (Line~\ref{algo_line:covariance_computation_start}--\ref{algo_line:covariance_computation_end}). In Line~\ref{algo_line:fif_computation_start}--\ref{algo_line:fif_computation_end}, we compute $ V_{\mathbf{a}, \mathbf{S}} $ for the most and the least favored groups, and plug them in Theorem~\ref{lemma:fif} to compute FIF of $ \nonsensitive_{\mathbf{S}} $.

@@

\paragraph{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
We extend {\fairXplainer} in computing FIF of group fairness metrics beyond statistical parity, namely equalized odds and predictive parity. For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of samples in the dataset where $ Y = 1 $ and another on samples with $ Y = 0 $. Then, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ quantifies the equalized odds of the classifer.
